Modelling BCG vaccination in the UK

### 2019-08-02 rmds/09-model-fitting.Rmd
model fitting comments from Ellen

The primary data source for the model is the reported, age-stratified, UK born TB notifications from 2000 to 2004 as recorded in the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). 2000 to 2004 are the years for which notifications are stratified by UK birth status with universal school-age BCG vaccination. Data was grouped using the age groups present in the dynamic model (5 year age groups up to 49 and then a group from 50-69 and a 70+) group. Using the age-stratified incidence data allows for more complex trends to be identified.

Additional datasets were considerd during initial model fitting and during model calibartion. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. Reducing the impact of stochastic noise, which makes fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were used as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. The downside of using this data is that twofold. Firstly, reporting standards may have changed over time so a single measurement model may not be appropriate. Secondly, non-UK born cases are included in these data making fitting to this data very dependent on

@@

There are three major considerations to account for when developing an observed disease notification model (i.e a reporting model). These are: systematic reporting error over time; systematic changes in reporting error over time; and reporting noise. For the notification data outlined in the previous section there is little evidence to suggest the form that this measurement model should take. For this reason I have assumed that all reporting errors are Gaussian and that their are no time variable reporting errors. The reporting model can be then defined as follows,

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that cases are less likely to be reported than they are to be over reported. The prior for the reporting noise is an assumption. This observation model was also used for the non-UK born pulmonary cases.

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000 (see Chapter \@ref(data)). A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, given that I have only included a single point prior to the introduction to the ETS doing so in this instance would lead to over-fitting of these parameters. Instead I have assumed that $ E_{\text{noise}}$ is doubled for data points prior to the introduction of the ETS.

@@

The model outlined in the Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful information in the observed data may be lost.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tols available. Ranging for tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates. Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to calibrate, and fit, the previously detailed state space model of TB transmission in England.

@@

1. For a given parameter set the particle filter is initialised by drawing a number of random samples (state particles) from the initial conditions of the model under consideration. These samples are then given a uniform weighting.

2. Sequentially for the each observed data point the particle filter is then advanced through a series of *propagation*, *weighting*, and *re-sampling* steps.

  * *Propagation*: For each particle the model is simulated, producing a forecast of the observed data data point.

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate Bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC). This algorithm is used for all dynamic model fitting in this thesis.

@@

Where $y$ is the observed data, $\theta$ is the current parameter set, $\theta'$ is some proposed parameter set sampled from some proposal distribution $q(\theta' | \theta)$.[@Murray2015] By construction samples drawn using this rule are ergodic to the posterior distribution. This means that after convergence samples drawn using this rule may be considered as samples from the posterior distribution.

@@

A simple approach to setting the proposal is to run a series of MCMC steps and then calculate the acceptance rate. Based on the acceptance rate the width of the proposal distributions can then be adapted. By repeating these steps multiple times a proposal distribution which gives an acceptance rate within the desired bounds can be arrived at. This adaption can either be independent for each parameter or dependent (taking into account empirical correlations). The `adapt_proposal` function, from the `rbi.helpers` R package,[@Funk:2019uw] implements this approach and is used in this model fitting pipeline. In many models parameters are likely to have strong correlations (i.e between UK and Non-UK born mixing rate and effective contact rate) and in these scenarios it is likely that a dependent strategy for adapting the proposal distribution will more efficiently explore the posterior distribution. However, the downside of adapting the proposal distribution using dependent methods is that the resulting proposal is highly complex, is computationally expensive to compute and may breakdown in some areas of the posterior distribution.

@@

I chose to use, manual, independent proposal adaption method for several reasons. Firstly, when developing this pipeline the approaches implemented in `rbi.helpers` produced multiple transient errors in other `rbi` and `rbi.helpers` code. Secondly, the resulting dependent proposal distribution was highly complex, slow to compute, and difficult to debug. Finally, for SMC-SMC efficient exploration of the proposal distribution is less important than when using MCMC alone as SMC-SMC is initialised with multiple samples from the prior distribution. This means that multiple local maximas can be efficiently explored regardless of the proposal distribution used. The MCMC rejuvenation step then serves to provide additional samples from these local maximas. Proposal adaption was only carried out for the main model scenario with all other scenarios using this proposal distribution.

@@

Where y are the data, $\displaystyle p(y|\theta)$ is the likelihood function and C is a constant. C cancels out when comparing different models and therefore does not need to be calculated.

@@

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC-SMC step. Parameter sensitivity measures such as PRCC must be calculated separately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

The full model fitting pipeline can be summarised as follows^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/fit_model.R]:

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvenation steps in all scenarios evaluated. This resulted in spuriously tight priors. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and indicative only.

@@

@@

```{r 09-prior-posterior, warning = FALSE, message = FALSE}
sum_prior_posterior %>% 
  set_names(., nm = . %>% 
              str_replace("or", "or (95\\\\%CI)")) %>%
  mutate(Parameter = latex_param) %>% 
  select(-latex_param) %>% 
kable(
  caption.short = "Prior and posterior medians for fitted model parameters", 
  caption = "Prior and posterior medians for fitted model parameters. (95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.",
  booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

>>>>

The primary data source for the model is the reported, age-stratified, UK born TB notifications from 2000 to 2004 as recorded in the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). 2000 to 2004 are the years for which notifications are stratified by UK birth status with universal school-age BCG vaccination. Data was grouped using the age groups present in the dynamic model (5 year age groups up to 49 and then a group from 50-69 and a 70+ group). Using the age-stratified incidence data allows for more complex trends to be identified.

Additional datasets were considerd during initial model fitting and duringl later model calibartion. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. Reducing the impact of stochastic noise, which makes fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were used as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. The downside of using this data is that twofold. Firstly, reporting standards may have changed over time so a single measurement model may not be appropriate. Secondly, non-UK born cases are included in these data making fitting to this data very dependent on

@@

There are three major considerations to account for when developing an observed disease notification model (i.e a reporting model). These are: systematic reporting error over time; systematic changes in reporting error over time; and reporting noise. I assumed that all reporting errors are Gaussian and that their are no time variable reporting errors. This model was used for all data fitted to. The reporting model can be defined as follows,

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that under reporting is more likely that over reporting. The prior for the reporting noise is based on the observed variation between years.

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000 (see Chapter \@ref(data)). A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, this may result in overfitting and so has not been implemented here.

@@

The model outlined in Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful information in the observed data may be lost.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates. Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

@@

1. For a given parameter set, the particle filter is initialised by drawing a number of random samples (state particles) from the initial conditions of the model under consideration. These samples are then given a uniform weighting.

2. Sequentially for the each observed data point, the particle filter is then advanced through a series of *propagation*, *weighting*, and *re-sampling* steps.

  * *Propagation*: For each particle the model is simulated, producing a forecast of the observed data point.

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models, a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate Bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC).[@Murray2015] This algorithm is used for all dynamic model fitting in this thesis.

@@

Where $y$ is the observed data, $\theta$ is the current parameter set, $\theta'$ is some proposed parameter set sampled from some proposal distribution $q(\theta' | \theta)$.[@Murray2015] By construction, samples drawn using this rule are ergodic to the posterior distribution. This means that after convergence, samples drawn using this rule may be considered as samples from the posterior distribution.

@@

A simple approach to setting the proposal is to run a series of MCMC steps and then calculate the acceptance rate. Based on the acceptance rate the width of the proposal distributions can then be adapted. By repeating these steps multiple times a proposal distribution which gives an acceptance rate within the desired bounds can be arrived at. This adaption can either be independent for each parameter or dependent (taking into account empirical correlations). The `adapt_proposal` function, from the `rbi.helpers` R package,[@Funk:2019uw] implements this approach and is used in this model fitting pipeline. In many models, parameters are likely to have strong correlations (i.e between UK and Non-UK born mixing rate and effective contact rate). In these scenarios, it is likely that a dependent strategy for adapting the proposal distribution will more efficiently explore the posterior distribution. However, the downside of adapting the proposal distribution using dependent methods is that the resulting proposal is highly complex, is computationally expensive to compute and may breakdown in some areas of the posterior distribution.

@@

I chose to use manual independent proposal adaption methods for several reasons. Firstly, when developing this pipeline the approaches implemented in `rbi.helpers` produced multiple transient errors in other `rbi` and `rbi.helpers` code. Secondly, the resulting dependent proposal distribution was highly complex, slow to compute, and difficult to debug. Finally, for SMC-SMC efficient exploration of the proposal distribution is less important than when using MCMC alone as SMC-SMC is initialised with multiple samples from the prior distribution. This means that multiple local maximas can be efficiently explored regardless of the proposal distribution used. The MCMC rejuvenation step then serves to provide additional samples from these local maximas. Proposal adaption was only carried out for the main model scenario with all other scenarios using this proposal distribution.

@@

Where y are the data, $\displaystyle p(y|\theta)$ is the likelihood function and $C$ is a constant. $C$ cancels out when comparing different models and therefore does not need to be calculated.

@@

I have implemented PRCC using the `epiR` R package^[Sensitivity code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC-SMC step. Parameter sensitivity measures such as PRCC must be calculated separately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Sensitivity plotting code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

The full model fitting pipeline can be summarised as follows^[Model fitting code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/fit_model.R]:

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvenation steps in all scenarios evaluated. This resulted in spuriously tight posterior distributions. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and indicative only.

@@

```{r 09-prior-posterior, warning = FALSE, message = FALSE}
sum_prior_posterior %>% 
  set_names(., nm = . %>% 
              str_replace("or", "or (95\\\\%CI)")) %>%
  mutate(Parameter = latex_param) %>% 
  select(-latex_param) %>% 
kable(
  caption.short = "Prior and posterior medians for fitted model parameters", 
  caption = "Prior and posterior medians for fitted model parameters. (95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.",
  booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>% 
  {.}
```

@@
